### 2011-10-20 doc/manuscript.tex
I am still merging Ilan's parts and mine.

\subsection*{Recombining gene transfer from SPY to SDE at genome-wide scale}
% \subsection*{Bayesian analysis of recombining gene transfer}

We followed the pipeline and methods recommended by Didelot et al.
\cite{Didelot2010} to pursue a model-based study of recombining gene transfer. A
preliminary step in the pipeline involved estimating the species
tree with branch lengths (via \texttt{ClonalFrame}), and three population
parameters of mutation rate, recombination rate, and average recombinant tract
length (via a first phase of \texttt{ClonalOrigin}).  Because 
\textit{S.\ equi} ssp.\textit{\ equi} strain 4047 (SEE)
diverged significantly from the others among the six genomes (Table S\ref{tab:genome}), we
could not obtain sufficiently long multiple sequence alignments with
it, and had to exclude it from the model-based analysis.  The estimated species tree
using the alignment of the five genomes (Figure
S\ref{fig:mauve}) had the expected topology \texttt{The figure 1 must be nicer
than one that now.} (see Figure \ref{fig:tree5}). \texttt{ClonalFrame} also
estimated
a relative strength of recombination compared to mutations ($r/m=4.66$), 
indicating potentially significant roles of recombination in the evolutionary
process of shaping these genomes; four times more sites had been affected by
recombination-driven substitutions than by mutation-driven substitutions.
($r/m$; Table S\ref{tab:clonalframe}).

Estimates of population mutation rate ($\theta$), recombination rate ($\rho$),
and average tract length ($\delta$) were obtained for the alignment blocks
analyzed (Figure S\ref{fig:scatter3}), and 
medians of each of the three parameters weighted over block lengths were
obtained as global estimates of the three population
parameters (Table S\ref{tab:three} and see \textbf{Materials and Methods}).  
The local estimates typically seemed to be
moderately fluctuating around the global average, with the exception of what
appeared to be a recombination hotspot at a region 200 kilobase pairs from the
origin of replication of SDE1.

The species tree and estimated branch lengths, together with the global
estimates of the three population parameters were used in the second phase of
ClonalOrigin, where recombinant trees were sampled from the posterior
distribution. One way to summarize the posterior sample of recombinant trees
provided by ClonalOrigin was to observe the topologies of the local genealogies
induced by it. Out of the 105 possible rooted labeled topologies spanning five
taxa, the dominant topology (sampled 66.7\% of the time) was that of the species
tree (Materials and Methods; Table S\ref{tab:Gene-tree-topologies}). The second
most frequently sampled topology (sampled 9.3\% of the time) was the one in
which SPY and SDE were sister clades (with SDD as an outgroup). Other topologies
were sampled with frequency less than 4\%. This gave initial indication to the
extent of recombining gene transfer  between SPY and SDE.

To gain a more fine-scale view of the main types of recombination scenarios
sampled by ClonalOrigin, we looked at the recombinant edges present in the
sampled recombinant trees. Following the approach of Didelot et al. \cite{Didelot2010}, we
classified these edges by types defined by distinct pairs of donor and recipient
branches of the species tree (Figure \ref{fig:tree5}). For each edge type, we recorded
the average number of edges of that type present in the sampled recombinant
tree, and computed the ratio of this observed count and the prior expected count
as implied by the species tree (considering branch lengths and their relative
positions) and the three
population parameters estimated in the preliminary round of the analysis. The
base 2 logarithms of these ratios are depicted in the heatmap of Figure
\ref{fig:Heatmap-of-recombination} (see also Tables
S\ref{tab:heatmap} and S\ref{tab:obsheatmap}).

Among the possible recombinant edge types, we distinguished between ones that
preserve the species tree topology, and ones that alter it. We focused
on the topology-altering types, both because our main objective was to study
cross-clade gene transfer, and because our simulation study indicates that the
observed counts of topology-preserving recombinant edges were 
unreliably inferred {\bf[IG: POINT TO RELEVANT SECTION]}. The heatmap allowed to find
recombination scenarios that deviated from the expectation dictated by the
probabilistic model of ClonalOrigin. {\bf [IG: CHECK THIS FOLLOWING LINE OF
ARGUMENT]} We envisioned two main potential causes for such a deviation: (1)
reduced or increased opportunities for species to interact and exchange genetic
material; (2) selection acting on newly transferred genes.

We observed a reduced count of recombinant edges sampled between SDD and the
three branches of the SPY clade ($\sim$25\% of the expected count). This is
consistent with the fact that these species have reduced opportunities to
interact with each other, SPY being a human pathogen and SDD being strictly veterinary. On the
contrary, the number of sampled recombinant edges between SDD and branches in
the SDE clade seemed to be consistent with the prior, with a slight excess for
edges from SDE1 and SDE2 to SDD {\bf [IG: IS THIS WORTH DWELLNG ON?]}. The
greatest deviation of observed counts from expected ones was observed for
recombinant edges between the SPY and SDE clades. It was concordant with the
fact that these two species share an ecological niche; SDE was an
opportunistic human pathogen. Interestingly, there appeared to be much higher
enrichment (up to four times more than expected) for recombinant edges
originating in the SPY clade and ending up in the SDE clade. In a follow-up
simulation study we show that this deviation from the expected count was indeed
highly significant, and so was the observed asymmetry in pattern of recombination
{\bf[IG: POINT TO RELEVANT SECTION]}. This bias in directionality of
recombination pattern between SPY and SDE was especially intriguing in light of
the hypothesis that SDE acquired human pathogenicity via gene transfer from SPY.

@@

{\bf[IG: DID NOT TOUCH THE FIRST 3 PARAGRAPHS. REQUIRES ``CONDENSING".]}
{\bf[IG: START WITH SOME GENERAL STATISTICS - NUMBER OF GENES, ETC?]}
\texttt{SCC: Check this ``condensed'' version.}

Figure \ref{fig:Gene-duplication-loss} shows gene gain, loss, duplication, and
two different gene transfer events along the lineages of the phylogenetic tree
that relates the six genomes. SPY had less genes than the other species. All of
the genomes were used to construct 2314 gene families.  The largest family
consisted of 59 genes while nearly half of gene families (1066) had exactly one
gene per species, or six genes total (Figure S\ref{fig:famsizes}).  We estimated
a larger number of recombining gene transfer events than that of horizontal gene
transfer events; 944 and 294, respectively.  Duplications in extant lineages
were more frequent than those in the other lineages.  SDE lineages from the root
appeared to have increased in the number of genes while the common ancestor of
SPY1 and SPY2 decreased in the number of genes.  Whereas gene loss events were
either comparable to or less than gene gain ones along most of the lineages, it
was not along two internal branches of SPY and SDE.  This indicated that SPY and
SDE lineages had reduced number of genes to specialize in their specific niches
after each split from their corresponding direct ancestors.

As in the Bayesian analysis of ClonalOrigin, we attempted to see which type of
horizontal or recombining transfer events was most common in our inference.  We
classified events into types defined by donor-recipient branches in the species
tree. We recorded the number of inferred events of each type in Figures
S\ref{fig:hgt-heatmap} and S\ref{fig:mowgli-recomb-heatmap}. Unlike in the
Bayesian analysis, we could not normalize these counts by their expectation
because we did not have any model-based expectation for them. This restricted
our ability to interprest these counts. We did notice, however, a large number
of inferred recombining gene transfer events from the ancestral SPY branch to
the ancestral SDE branch, consistent with our observations from the Bayesian
analysis. With horizontal gene transfer, we did not observed a similar pattern,
but we did observe a large number of horizontal transfers from the ancesral SDE
branch to SDD.

@@

We set out to seek categories of genes that were significantly enriched for
recombining or horizontal transfer events. For horizontal
transfer, we used a score based on the number of events inferred by the
parsinomy-based method, and for recombining gene transfer, we used both the
parsimony-based inferred counts and a recombination intensity score based on the
Bayesian analysis (see \textbf{Materials and Methods}). First, we checked association of
transfers with virulence genes, which were provided from a previous study
of Suzuki et al.\ \citet{Suzuki2011}.
This set of virulence genes was found to be significantly associated
with horizontal gene transfers (P-value $<1.33 \times 10^{-11}$), but not with
recombining gene transfers. We also found a marginally
significant association between virulence genes and gene duplications (P-value
$<7.13 \times 10^{-3}$).

@@

\subsection*{Comparison of Bayesian and parsimony-based inferences of recombining gene transfer}

The similarity between the two lists of gene categories found to be associated
with recombining gene transfers in the separate analyses -- Bayesian and
parsimony-based --  indirectly indicated concordance between the two independent
approaches.  In order to obtain a more direct measure of correlation, we used
the set of genes present in both analyses to compute the correlation between
recombination intensity scores and the number of recombining gene transfers
inferred by the parsimony-based approach. We observed a positive correlation of
$0.39$ (P-value $2.2\times10^{-16}$; Figure S\ref{fig:cmpcomowgli}) between the
two measures.  For comparison, the correlation between recombination intensity
and the number of horizontal gene transfers was 0.03 (P-value 0.18;
S\ref{fig:cmpcomowglihgt}).

@@

Often computational and statistical analyses on genomes could benefit other
researchers when those resources were provided via Genome Browser.  We wished to
provide research community especially in SPY and SDE bacteria with results from
our analysis of gene transfer.  The resource shown in Figure S\ref{fig:ucsc} for
gene transfer is available at \url{http://strep-genome.bscb.cornell.edu}.  For
each of the five individual genomes as a reference genome several annotation
tracks were displayed: alignment blocks, recombination rate per site for each
block, local gene tree topologies, recombination intensity, locations of
virulence genes, posterior probability of recombining gene transfer. The genome
track of probability of recombining gene transfer allowed us to easily browse
the bacterial genome to locate regions that we might be interested in.  Using
the genome browser tracks we located virulence genes with notable posterior
recombination probability (Table S\ref{tab:virrecomb}).  About 10\% of the genes
comprised of virulence facters, about half of which were found in the core
genome of the five genome alignment \citep{Suzuki2011}.  By sorting virulence
genes by the number of horizontal gene transfer events we listed several genes
with high ranks (Table S\ref{tab:virhgt}).
% Events counts for genes.

The process of gene transfer may or may not move whole genes from one species to
another \citep{Chan2009}. We observed that units of recombining gene
transfer could be multiple genes (Table S\ref{tab:genes-transfer}). In some
cases, parts of a gene were inferred to have been transferred, for instance,
transfer from SPY1 to SDE1 was inferred for 29\% of the aspartyl-tRNA synthetase
gene. Recombinant tracts often were shown to span two genes.  In other cases,
the inferred transfer units spaned a set of as much as eight adjacent genes:
loci from SpyM3\_1518 to SpyM3\_1525.

\subsection{Simulations}

@@

\section*{Discussions}

Molecular evolution of bacterial genomes is multifaceted. Studies of bacterial
evolution often focused on one side of many evolutionary processes.  While
understanding one aspect of bacterial evolutionary processes we could be
ignorant of the other features.  Horizontal gene transfer process was often
considered in macro-evolutionary scale using a great deal of bacterial species.
As closely related species or groups of bacterial individuals are becoming a
focus of evolutionary study, recombining gene transfer process would surface as
another facet of evolutionary forces.  Because the two gene transfer processes
would act on bacterial genomes simultaneously, ignoring one of the two processes
in studying evolution of bacteria would lead to a narrowed point of view to the
underlying evolutionary forces. Our study of SPY and SDE genomes undertook the
work of discriminating recombining and non-recombining horizontal gene transfer
processes because the two species were very closely related.  We showed that the
two gene transfer processes acted on different biological functions. Recombining
gene transfer process in the analysis of SPY and SDE genomes was related with
genes encoding membrane-bound molecules and aminoacyl tRNA synthetases. On the
contrary, horizontal gene transfer process was found to be associated with
transposases, which would allow horizontal gene transfer of their neighboring
genes.  Because SPY was a human-specific pathogen, and SDE was an opportunistic
human pathogen, we were interested in association of pathogenicity with the two
gene transfer processes. We showed that virulence genes were more attributed to
horizontal gene transfer process than to recombining gene transfer process.
With the current data of only SPY and SDE genomes we could not conclude that the
association of virulence genes with horizontal gene transfer not recombining
gene transfer was applicable to other species.  However, without the
discrimination of the two processes it would have been difficult to find the
disparate association of biological functions or virulence genes with the
different gene transfer processes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsection*{Distinguishing two gene transfer processes shows different functional
categories associations}

Bacterial gene transfer processes as an evolutionary force would affect more
frequently in some biological functions than in others.  In our attempt of
finding which biological functions were associated with recombining or
horizontal gene transfers, we devised a measure of recombination intensity for
genes by using frequency of recombinant edges that affect those genes, and used
frequency of horizontal gene transfer events (see section \textbf{Materials and
Methods}).  Table \ref{tab:functional} shows a list of function categories
that were found to be associated with recombining gene transfers. We noted three
biological functions in the list: membrane-bound molecules, aminoacyl tRNA
synthetases, and ribosomes.

Firstly, we have sugar phosphotransferases ranked
top in the list; phosphrylation could be important in signal transduction to
recognize environmental changes.  Transport functional categories of
carbohydrate transport and sodium ion transport would belong to genes encoding
membrane-bound molecules.  Also highly ranked were fatty acid and lipid
biosyntheses that must be involved in producing membrane-bound molecules. 
Because bacteria as a unicellular organism should be able to respond to changes
to environment, some evolutionary pressures must have been acting on genes
encoding membrane-bound molecules. Therefoure, recombining gene transfer process
appeared to be more associated with biological functions that had been critical
of sensing queues of bacterial environmental changes.

Secondly, we found several aminoacyl tRNA synthetases (GO Term translation
GO:0006412) in association with recombining gene transfer.  In the
parsimony-based approach the term translation was also top ranked (Table
S\ref{tab:go-events-recombining}).  Because they had been a focus of gene
transfer in bacterial species \citep{Woese2000}, we could confirmed that the
recombination intensity and requency of recombining gene transfer events would
be useful in finding genes associated with recombining gene transfer.

Thirdly, functional categories concerning ribosomes were unexpectedly
found to be associated with recombining gene transfer.  Because genes related
with ribosomes were considered to be housekeeping genes, gene transfer in which
would be rare, it was unexpected that ribosome related genes were found to be
associated with recombining gene transfer.  Because it was a bizarre result, we
investigated the association by considering different recombinant edge types. We
noted that two different classes of recombinant edge types: one that changes
species tree topologies, and another that does not
(see section \textbf{Materials and Methods}).
Because recombinant edges
that did not change species tree topology might be difficult to infer reliably,
considering those recombinant edges in computing recombination intensity might
lead to dubious results.  Ribosome related functional categories were found to
be top ranked if we took those recombinant edges that did not change species
tree topology (Tables S\ref{tab:functional-notopology} and
S\ref{tab:functional-all}).  In Table \ref{tab:functional} ribosome related
functional categories were not top ranked.  Therefore, high recombination
intensity in ribosome related genes were likely to be attributed to
topology-not-changing recombinant edges. We would be cautious of recombining
transfer of genes related with ribosome related functions.

Table \ref{tab:go-events} shows a list of functional categories associated with
horizontal gene transfer, which were different from those from recombining gene
transfer analysis.  Ranked top in the list was transposases, which were known to
help foreign genes be excised from a donor genome and integrated into a
recipient genome. The biological role of transposases was concordant with
horizontal gene transfer process because the process involved in transposases
would allow bacteria to acquire foreign genes without the requirement of
homologous regions in the bacteria.  
These differences in the profile of functional categories association of the two
gene transfer processes were because of the approach of dicriminating the two
gene transfer processes.

% The above was moved from Results of functional category analysis.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

{\bf [IG: MOVE TO DISCUSSION:]}
A full model-based approach would be desirable in analyzing data for testing
relevant hypotheses more systematically. Model complexity might not allow easy
access to the implementation of a model. Our study involved a model-based
approach and a parsimony-based one for understanding evolution in the core
genomes of several bacteria individuals. Because of inability of the model-based
method for studying the pan genomes we had to resort to the parsimony-based
method. However, the parsimony-based method was also applied to the core
genomes. We wished to compare the two methods in their results of the core
genome analyses.  Because of two independent analyses on the same data set, this
would allow us to be more confident of the results if the two analyses performed
similarly Generally, phylogenetic methods of detecting gene transfers often take
advantage of discrepancy between a reference species tree and gene trees. The
two approaches that we employed in this study were no exception. 
Another contribution of our study to bacterial gene transfer research is to
elucidate the net directionality of gene flow from SPY to SDE in the core
genomes. Although the issue of net directionality has been tackled, our study
was the first attempt of resolving the issue in genome-wide scale. Yet, it still
remains to be seen whether this net directionality of gene flow from SPY to SDE
happened in the pan genome. A model-based method with more refined models would
be necessary for the reserach direction because the parsimony-based method we
employed simply could not show any net directionality between SPY and SDE in the
pan genome. Although we are unsure whether much larger number of genomes could
be useful in finding net directionality, we consider that more SDE genomes would
be beneficial to future studies. 
%
% The above was moved from Results of functional category analysis.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

@@

%=============================================================================
% Table S. Mowgli - Recombining Transfer
\begin{table}
\caption{\label{tab:go-events-recombining}Functional categories enriched in 
recombining transfer events that are inferred using the parsimony-based approach.}
\noindent \begin{centering}
\begin{tabular}{cccl}
\hline 
P-value & Count & GO-term & Description \\
\hline 
7e-09 & 153 & GO:0006412 & translation\\
1e-08 &  67 & GO:0030529 & ribonucleoprotein complex\\
1e-07 &  45 & GO:0019843 & rRNA binding\\
2e-06 &  73 & GO:0003735 & structural constituent of ribosome\\
3e-05 & 157 & GO:0005840 & ribosome\\
7e-05 & 221 & GO:0003723 & RNA binding\\
2e-04 &  11 & GO:0004826 & phenylalanine-tRNA ligase activity\\
3e-04 &  14 & GO:0015413 & nickel-transporting ATPase activity\\
4e-04 &  93 & GO:0008982 & protein-N(PI)-phosphohistidine-sugar phosphotransferase activity\\
5e-04 & 104 & GO:0009401 & phosphoenolpyruvate-dependent sugar phosphotransferase system\\
\hline 
\end{tabular}
\par\end{centering}
\end{table}
%\clearpage{}
%=============================================================================

@@

Virulence genes encoding putative
nucleases with a secretion signal peptide were also found to have had
experienced horizontal gene transfer:

>>>>

\subsection*{Recombining gene transfer from SPY to SDE at genome-wide scale}

We followed the pipeline and methods recommended by Didelot et al.
\cite{Didelot2010} to pursue a model-based study of recombining gene transfer. A
preliminary step in the pipeline involved estimating the species tree with
branch lengths (via \texttt{ClonalFrame}), and three population parameters of
mutation rate, recombination rate, and average recombinant tract length (via a
first phase of \texttt{ClonalOrigin}).  Because \textit{S.\ equi} ssp.\textit{\
equi} strain 4047 (SEE) diverged significantly from the others among the six
genomes (Table S\ref{tab:genome}), we could not obtain sufficiently long
multiple sequence alignments with it, and had to exclude it from the model-based
analysis.  The estimated species tree using the alignment of the five genomes
(Figure S\ref{fig:mauve}) had the expected topology \texttt{The figure 1 must be
nicer than one that now} (see Figure \ref{fig:tree5}). \texttt{ClonalFrame} also
estimated a relative strength of recombination compared to mutations
($r/m=4.66$), indicating potentially significant roles of recombination in the
evolutionary process of shaping these genomes; four times more sites had been
affected by recombination-driven substitutions than by mutation-driven
substitutions.  (Table S\ref{tab:clonalframe}).  Estimates of population
mutation rate ($\theta$), recombination rate ($\rho$), and average tract length
($\delta$) were obtained for the alignment blocks analyzed (Figure
S\ref{fig:scatter3}), and the median weighted over block lengths for each of the
three parameters was obtained as a global estimate (Table S\ref{tab:three} and
see \textbf{Materials and Methods}).  The local estimates typically seemed to be
moderately fluctuating around the global average, with the exception of what
appeared to be a recombination hotspot at a region 200 kilobase pairs from the
origin of replication of SDE1.

The estimated species tree together with the global estimates of the three
population parameters were used in the second phase of \texttt{ClonalOrigin},
where recombinant trees were sampled from the posterior distribution. One way to
summarize the posterior sample of recombinant trees was to count the topologies
of the local genealogies induced by it. Of the 105 possible rooted topologies
relatig five taxa, the dominant topology (66.7\%) was that of the species tree
(see \textbf{Materials and Methods} and Table S\ref{tab:Gene-tree-topologies}).
The second most frequently sampled topology (9.3\%) was the one in which SPY and
SDE were sister clades to each other with SDD being as an outgroup. Other
topologies were sampled with frequency less than 4\%. This gave initial
indication to the extent of recombining gene transfer  between SPY and SDE.

To gain a more fine-scale view of the main types of recombination scenarios, we
looked at the recombinant edges present in the posterior sample of recombinant
trees. Following the approach of Didelot et al. \cite{Didelot2010}, we
classified these edges by types defined by distinct pairs of donor and recipient
branches of the species tree (Figure \ref{fig:tree5}). For each edge type, we
recorded the average number of edges of that type present in the sampled
recombinant tree, and computed the ratio of this observed frequency to the prior
expected one as implied by the species tree (considering branch lengths and
their relative positions) and the three population parameters estimated in the
preliminary step of the model-based analysis.  Base 2 logarithms of these ratios
are depicted in the heatmap of Figure \ref{fig:Heatmap-of-recombination} (see
also Tables S\ref{tab:heatmap} and S\ref{tab:obsheatmap}).

Among the possible recombinant edge types, we distinguished between ones that
preserve the species tree topology, and ones that alter it. We focused on the
topology-altering types, both because our main objective was to study
cross-clade gene transfer, and because our simulation study indicated that the
observed counts of topology-preserving recombinant edges were unreliably
inferred (see \textbf{Simulations}). The heatmap allowed to find recombination
scenarios that deviated from the expectation dictated by the probabilistic
model.  We envisioned two main potential causes for such deviations: (1) reduced
or increased opportunities for species to interact and exchange genetic
material; (2) selection acting on newly transferred genes.

We observed a reduced count of recombinant edges sampled between SDD and the
three branches of the SPY clade ($\sim$25\% of the expected count). This was
consistent with the fact that these species had few opportunities to interact
with each other, SPY being a human pathogen and SDD being strictly veterinary.
On the contrary, the number of sampled recombinant edges between SDD and
branches in the SDE clade seemed to be consistent with the prior, with a slight
excess for edges from SDE1 and SDE2 to SDD.  The greatest deviation of observed
counts from expected ones was observed for recombinant edges between the SPY and
SDE clades. It was concordant with the fact that these two species shared an
ecological niche; SDE was an opportunistic human pathogen. Interestingly, there
appeared to be much higher enrichment (up to four times more than expected) for
recombinant edges originating in the SPY clade and ending up in the SDE clade.
In a follow-up simulation study we show that this deviation from the expected
count was indeed highly significant, and so was the observed asymmetry in
pattern of recombination (see \textbf{Simulations}).  This net directionality of
recombination pattern from SPY to SDE was especially intriguing in light of the
hypothesis that SDE acquired human pathogenicity via gene transfer from SPY.

@@

Figure \ref{fig:Gene-duplication-loss} shows the inferred numbers of gene gain,
loss, duplication, and two different kinds of gene transfer events along
lineages of the tree that relates the six genomes. SPY had less genes than the
other species. All of the genomes were used to construct 2314 gene families.
The largest family consisted of 59 genes while nearly half of gene families
(1066) had exactly one gene per species, or six genes total (Figure
S\ref{fig:famsizes}).  We estimated a larger number of recombining gene transfer
events than that of horizontal gene transfer events; 944 and 294, respectively.
Duplications in extant lineages were more frequent than those in the other
lineages.  SDE lineages from the root appeared to have increased in the number
of genes while the common ancestor of SPY1 and SPY2 decreased in the number of
genes.  Whereas gene loss events were either comparable to or less than gene
gain ones along most of the lineages, it was not the case along the two internal
branches of SPY and SDE.  This indicated that SPY and SDE lineages had reduced
number of genes to specialize in their specific niches after each divergence
from their corresponding direct ancestors.

As in the model-based analysis, we attempted to see which type of horizontal or
recombining transfer events was most common in our inference.  We classified
events into types defined by donor-recipient branches in the species tree. We
recorded the number of inferred events of each type (Figures
S\ref{fig:hgt-heatmap} and S\ref{fig:mowgli-recomb-heatmap}). Unlike in the
model-based analysis, we could not normalize these counts by their expectation
because we did not have any model-based expectation for them, which restricted
our ability to interprest these counts. We did notice, however, a large number
of inferred recombining gene transfer events from the ancestral SPY branch to
the ancestral SDE branch, consistent with our observations from the model-based
analysis. Yet, we did not observed a similar pattern with horizontal gene
transfer while we observed a large number of horizontal transfers from the
ancesral SDE branch to SDD.

@@

We set out to seek categories of genes that were significantly enriched for
recombining or horizontal transfer events. For horizontal transfer, we used a
score based on a number of events inferred by the parsimony-based method, and
for recombining gene transfer, we used both the parsimony-based inferred counts
and a recombination intensity score based on the model-based analysis (see
\textbf{Materials and Methods}). First, we checked association of transfers with
virulence genes, which were provided from a previous study of Suzuki et al.\
\citet{Suzuki2011}.  This set of virulence genes was found to be significantly
associated with horizontal gene transfers (P-value $<1.33 \times 10^{-11}$), but
not with recombining gene transfers. We also found a marginally significant
association between virulence genes and gene duplications (P-value $<7.13 \times
10^{-3}$).

@@

\subsection*{Comparison of model-based and parsimony-based inferences of recombining gene transfer}

The similarity between the two lists of gene categories found to be associated
with recombining gene transfers in the separate analyses -- model-based and
parsimony-based --  indirectly indicated concordance between the two independent
approaches.  In order to obtain a more direct measure of correlation, we used
the set of genes present in both analyses to compute the correlation between
recombination intensity scores and the number of recombining gene transfers
inferred by the parsimony-based approach. We observed a positive correlation of
$0.39$ (P-value $2.2\times10^{-16}$; Figure S\ref{fig:cmpcomowgli}) between the
two measures.  For comparison, the correlation between recombination intensity
and the number of horizontal gene transfers was 0.03 (P-value 0.18;
Figure S\ref{fig:cmpcomowglihgt}).

@@

Often computational and statistical analyses on genomes could benefit other
researchers when those resources were provided via Genome Browser.  We wished to
provide research community especially in SPY and SDE bacteria with results from
our analysis of gene transfer.  The resource shown in Figure S\ref{fig:ucsc} for
gene transfer was available at \url{http://strep-genome.bscb.cornell.edu}.  For
each of the five individual genomes as a reference genome several annotation
tracks were displayed: alignment blocks, recombination rate per site for each
block, local gene tree topologies, recombination intensity, locations of
virulence genes, posterior probability of recombining gene transfer. The genome
track of probability of recombining gene transfer allowed us to easily browse
the bacterial genome to locate regions that we might be interested in.  Using
the genome browser tracks we located virulence genes with notable posterior
recombination probability (Table S\ref{tab:virrecomb}).  About 10\% of the genes
comprised of virulence facters, about half of which were found in the core
genome of the five genome alignment \citep{Suzuki2011}.  By sorting virulence
genes by the number of horizontal gene transfer events we listed several genes
with high ranks (Table S\ref{tab:virhgt}).
The whole gene of  putative dipeptidase (SpyM3\_0465) was transferred, which was
a case for the finding that virulence genes tended to be transferred as a whole
unit \citep{Chan2009}.  Another example of whole gene transfer was superoxide
dismutase (SpyM3\_1071).  We investigated the unit of recombining gene transfer
by finding 58 genes for which a recombinant edge of certain types were sampled
with probability at least 0.9 by focusing on specific types of recombinant edges
(Table S\ref{tab:genes-transfer}).  In some cases, parts of a gene were inferred
to have been transferred, for instance, transfer from SPY1 to SDE1 was inferred
for 29\% of the aspartyl-tRNA synthetase gene. Recombinant tracts often were
shown to span two genes.  In other cases, the inferred transfer units spaned a
set of as much as eight adjacent genes: loci from SpyM3\_1518 to SpyM3\_1525.

@@

\section*{Discussion}

Molecular evolution of bacterial genomes is multifaceted. Studies of bacterial
evolution often focused on one side of many evolutionary processes.  While
understanding one aspect of bacterial evolutionary processes we could be
ignorant of the other features.  Horizontal gene transfer process was often
considered in macro-evolutionary scale using a great deal of bacterial species.
As closely related species or groups of bacterial individuals are becoming a
focus of evolutionary study, recombining gene transfer process would surface as
another facet of evolutionary forces.  Because the two gene transfer processes
would act on closely related bacterial genomes simultaneously, ignoring one of the two processes
in studying evolution of bacteria would lead to a narrowed point of view to the
underlying evolutionary forces. Our study of SPY and SDE genomes undertook the
work of discriminating recombining and horizontal gene transfer
processes because the two species were very closely related.  We showed that the
two gene transfer processes acted on different biological functions. 
Recombining
gene transfer process in the analysis of SPY and SDE genomes was related with
genes encoding membrane-bound molecules and aminoacyl tRNA synthetases. On the
contrary, horizontal gene transfer process was found to be associated with
transposases, which would allow horizontal gene transfer of their neighboring
genes.  Because SPY was a human-specific pathogen, and SDE was an opportunistic
human pathogen, we were interested in association of pathogenicity with the two
gene transfer processes. We showed that virulence genes were more associated
with 
horizontal gene transfer process than with recombining gene transfer process.
With the current data of only SPY and SDE genomes we could not conclude that the
association of virulence genes with horizontal gene transfer not recombining
gene transfer was applicable to other species.  However, without the
discrimination of the two processes it would have been difficult to find the
disparate association of biological functions or virulence genes with the
different gene transfer processes.

Bacterial gene transfer processes as an evolutionary force would affect more
frequently in some biological functions than in others.  In our attempt of
finding which biological functions were associated with recombining or
horizontal gene transfers, we devised a measure of recombination intensity for
genes by using frequency of recombinant edges that affect those genes, and used
frequency of horizontal gene transfer events.  This approach allowed us to be
able to associate gene transfer with biological functions including
membrane-bound molecules, aminoacyl tRNA synthetases, and ribosomes.  Sugar
phosphotransferases could be important in signal transduction to recognize
environmental changes.  Transporters, and fatty acid and lipid biosyntheses
would be involved in membrane-bound molecules.  Because bacteria as a
unicellular organism should be able to respond to changes to environment, some
evolutionary pressures must have been acting on genes encoding membrane-bound
molecules. Therefoure, recombining gene transfer process appeared to be more
associated with biological functions that had been critical of sensing queues of
bacterial environmental changes.  Because several aminoacyl tRNA synthetases
belonging to gene ontology term of translation (GO:0006412) had been a focus of
gene transfer in bacterial species \citep{Woese2000}, we could confirm that the
recombination intensity and requency of recombining gene transfer events would
be useful in finding genes associated with recombining gene transfer.
The biological role of transposases was concordant with horizontal gene transfer
process because the process involved in transposases would allow bacteria to
acquire foreign genes without the requirement of homologous regions in the
bacteria.  These differences in the profile of functional categories association
of the two gene transfer processes were because of the approach of dicriminating
the two gene transfer processes.

Functional categories concerning ribosomes were unexpectedly found to be
associated with recombining gene transfer.  Because genes related with ribosomes
were considered to be housekeeping genes, gene transfer in which would be rare,
it was unexpected that ribosome related genes were found to be associated with
recombining gene transfer.  Because it was a bizarre result, we investigated the
association by considering different recombinant edge types.  Ribosome related
functional categories were found to be top ranked if we took
tree-topology-preserving recombinant edges into account (Tables
S\ref{tab:functional-notopology} and S\ref{tab:functional-all}).  whereas
ribosome related functional categories were not top ranked in Table
\ref{tab:functional}.  Therefore, high recombination intensity in ribosome
related genes were likely to be attributed to topology-preserving recombinant
edges that were found to be difficult to infer reliably. So, we would have to be
cautious of associating ribosome related functions with recombining gene
transfer.

Another contribution of our study to bacterial gene transfer research is to
elucidate the net directionality of gene flow from SPY to SDE in the core
genomes. Although the issue of net directionality has been tackled, our study
was the first attempt of resolving the issue in genome-wide scale. Yet, it still
remains to be seen whether this net directionality of gene flow from SPY to SDE
happened in the pan genome. A model-based method with more refined models would
be necessary for the reserach direction because the parsimony-based method we
employed simply could not show any net directionality between SPY and SDE in the
pan genome. Although we are unsure whether much larger number of genomes could
be useful in finding net directionality, we consider that more SDE genomes would
be beneficial to future studies.

@@

A full model-based approach would be desirable in analyzing data for testing
relevant hypotheses more systematically. Model complexity might not allow easy
access to the implementation of a model. Our study involved a model-based
approach and a parsimony-based one for understanding evolution in the core
genomes of several bacteria individuals. Because of inability of the model-based
method for studying the pan genomes we had to resort to the parsimony-based
method. However, the parsimony-based method was also applied to the core
genomes. We wished to compare the two methods in their results of the core
genome analyses.  Because of two independent analyses on the same data set, this
would allow us to be more confident of the results if the two analyses performed
similarly Generally, phylogenetic methods of detecting gene transfers often take
advantage of discrepancy between a reference species tree and gene trees. The
two approaches that we employed in this study were no exception.

@@

%=============================================================================
% Table S. Mowgli - Recombining Transfer
\begin{table}
\caption{\label{tab:go-events-recombining}
\bf{Functional categories enriched in 
recombining transfer events that are inferred using the parsimony-based
approach.}}
\noindent \begin{centering}
\begin{tabular}{cccl}
\hline 
P-value & Count & GO-term & Description \\
\hline 
7e-09 & 153 & GO:0006412 & translation\\
1e-08 &  67 & GO:0030529 & ribonucleoprotein complex\\
1e-07 &  45 & GO:0019843 & rRNA binding\\
2e-06 &  73 & GO:0003735 & structural constituent of ribosome\\
3e-05 & 157 & GO:0005840 & ribosome\\
7e-05 & 221 & GO:0003723 & RNA binding\\
2e-04 &  11 & GO:0004826 & phenylalanine-tRNA ligase activity\\
3e-04 &  14 & GO:0015413 & nickel-transporting ATPase activity\\
4e-04 &  93 & GO:0008982 & protein-N(PI)-phosphohistidine-sugar phosphotransferase activity\\
5e-04 & 104 & GO:0009401 & phosphoenolpyruvate-dependent sugar phosphotransferase system\\
\hline 
\end{tabular}
\par\end{centering}
\end{table}
%\clearpage{}
%=============================================================================

@@

@@
